Staged Abstract Interpreters

### 2019-03-15 paper/10_intro.tex
some revision

where $e$ is the program of type @Expr@ and @arg@ is the input to program $e$,
and @eval@ produces a value.
Semantically, the interpreter satisfies $ \texttt{eval}(e)(arg) = [\![ e ]\!] arg$ for all
programs and inputs. Given a program $e_0$, by applying the first Futamura
projection, we may obtain a specialized interpreter
$\texttt{eval}_{\texttt{e0}}$ with respect to $e_0$, and by definition of the interpreter, 
$\texttt{eval}_{\texttt{e0}}(arg) = [\![ e_0 ]\!] arg $. If we generalize the
argument @arg@ to an environment (as well as a companion heap object, if for the
interest of using assignments) that contains values mapped from free variables
in $e$, the interpreter is a standard environment/store-passing interpreter,
where the specialization through MSP is also applicable.

@@

In this paper, we study the confluence of two old ideas --- Futamura projection
and abstract interpretation, but from the perspective of their recent
realizations --- multi-stage programming and definitional abstract interpreters.
To be specific, we present the application of the first Futamura projection on
definitional abstract interpreters, which intersects as staged abstract
interpreters. By borrowing the ideas from monadic interpreters
\cite{Steele:1994:BIC:174675.178068, DBLP:conf/popl/LiangHJ95,
DBLP:journals/pacmpl/DaraisLNH17, Sergey:2013:MAI:2491956.2491979} and embedded
domain-specific languages \cite{DBLP:conf/snapl/RompfBLSJAOSKDK15,
DBLP:journals/jfp/CaretteKS09, DBLP:conf/icfp/GibbonsW14,
Hofer:2008:PED:1449913.1449935}, we develop an approach to fulfill four
different semantics (Figure~\ref{confluence}) in a unified framework. The four
semantics share one same generic interpreter interface. Later, by instantiating
the interpreter differently, it can behave as a concrete interpreter, an
abstracter interpreter, or their code-generator versions, respectively. The key
idea that enables this flexibility is to use abstract type members abstract over
concrete/abstract components (e.g., concrete values or abstract values), as well
as the binding-time of them (e.g., static or dynamic). By staging, the overhead
caused by the monadic layers is eliminated in the generated code.

@@

Besides the intellectual merit shown in the confluence diagram, this paper
contributes to a practical problem in static analysis: \textit{can we speed up a
static analyzer without any intrusive changes and soundness compromises?} Static
analysis is known as a trade-off game between \textit{precision} and
\textit{efficiency}, we argue that applying meta-programming (multi-stage
programming, in particular) with abstract interpreters is an effective approach
to improve the latter one while the former one is untouched. Specialization
removes the overhead of abstractions introduced by high-level programming (e.g.,
monads), but not loses any advantages of this programming style (e.g.,
extensibility), nor changes the meaning of the program -- in this case, the
abstract semantics that does the analysis. Indeed, specializing analysis
with respect to a program by partial evaluation \cite{damian1999partial,
amtoft1999partial, Boucher:1996:ACN:647473.727587, ashley:practical} is not a
new idea, despite the fact that most of the previous work are ad-hoc on a
particular analysis or requiring significant changes to the analysis. For
example, abstract compilation \cite{Boucher:1996:ACN:647473.727587} requires the
whole analyzer to be rewritten as closure generation from. In this paper, we
show a systematic approach to meta-programming that achieves the performance
goal with less intrusive changes; in fact, a case study (Section~\ref{cs_ac})
comparing with abstract compilation shows, by utilizing type-based stage
annotations, the analyzer program does not need any changes.

@@

Our second contribution is to enable modular analysis based on a whole-program
analyzer by meta-programming. Modern software are shipped with large libraries
\cite{toman_et_al:LIPIcs:2017:7121}, and analyzing programs with large libraries
is usually expensive or unnecessarily repeated for whole-program analyzing
algorithms. For example, it has been shown that analyzing a simple ``Hello
World'' program in Java depends on additional 3,000 classes in the library
\cite{DBLP:conf/oopsla/KulkarniMZN16}. A natural idea is to analyze libraries
and generate summaries, later we can plugin the necessary information into the
summaries and finish the analysis; therefore such summaries should be modular
and reusable. Meta-programming gives us a chance to readily tweak a
whole-program analyzing algorithm into an analyzer that works modularly and
produces summaries. Still, the idea is to specialize the abstract interpreter
that was used for whole-program analysis, but on a subset of the program, for
examples, several generic functions for list manipulation. Such specialization
works on lambda expressions and generates code, which is partial analysis
results and can be completed by running with the necessary arguments (for
example, an environment that closes the lambda term) at next stages. Multiple
partial analysis results also can be reused later and composed together as they
are just normal programs at the next stage. Through staging, we mechanically
obtain a modular analysis, even though the original algorithm is formulated as a
whole-program analysis.

@@

\paragraph{Contributions and Outline}

\begin{itemize}
  \item We progressively present a unifying construction that naturally extends
    the first Futamura projection to abstract interpreters. Starting from a generic
    interface of the interpreter for a small functional language
    (Section~\ref{generic_if}), we first show the instantiation of concrete
    interpretation (Section~\ref{unstaged_conc}). Then we lift it to the staged
    version by replacing the binding-time type (Section~\ref{stagedinterp}); and to
    the abstract interpreter by applying abstraction to the environment, store, and
    values (Section~\ref{unstaged_abs}). Finally, we show the combination of the two
    transformations, dubbed \textit{staged abstract interpreters}, can be easily
    derived (Section~\ref{sai}).

>>>>

where $e$ is the program of type @Expr@ and @arg@ is the input to program $e$,
and @eval@ produces a value.
Semantically, the interpreter satisfies $ \texttt{eval}(e)(arg) = [\![ e ]\!] arg$ for all
programs and inputs. Given a program $e_0$, by applying the first Futamura
projection, we may obtain a specialized interpreter
$\texttt{eval}_{\texttt{e0}}$ with respect to $e_0$ -- the so-called \textit{equivalent executable}. By definition of the interpreter, 
$\texttt{eval}_{\texttt{e0}}(arg) = [\![ e_0 ]\!] arg $. If we generalize the
argument @arg@ to an environment (as well as a companion heap object, if for the
interest of using assignments) that maps free variables in $e$ to some values, 
the interpreter is a standard environment/store-passing interpreter,
where the specialization through MSP is also applicable.

@@

In this paper, we study the confluence of two old ideas --- Futamura projection
and abstract interpretation, but from the perspective of their recent
realizations --- multi-stage programming and definitional abstract interpreters.
To be specific, we present the application of the first Futamura projection on
definitional abstract interpreters, which intersects as staged abstract
interpreters. We develop a unified approach to fulfill the four different
semantics, as shown in Figure~\ref{confluence}. Starting with a generic but
parametric interpreter shared by the four semantics, by simply instantiating
the interpreter differently, we can automatically derive a concrete interpreter,
or an abstract interpreter, or their code-generator versions, respectively.

\iffalse
embedded domain-specific languages \cite{DBLP:conf/snapl/RompfBLSJAOSKDK15,
DBLP:journals/jfp/CaretteKS09, DBLP:conf/icfp/GibbonsW14,
Hofer:2008:PED:1449913.1449935},
The key idea that enables this flexibility is to use abstract type members first
abstract over concrete/abstract components (e.g., concrete values or abstract
values), then as well as the binding-time of them (e.g., static or dynamic). By
staging, the overhead caused by the monadic layers is eliminated in the
generated code.
\fi

\paragraph{Abstraction, Abstraction and Abstraction}
To enable such flexibility between different semantics, we extensively use
abstract type members over multiple abstractions, which can be later instantiated.

\begin{itemize}
\item Monadic abstractions, which encapsulate the following two abstractions.
  The interpreter itself is written in monadic form
  \cite{Steele:1994:BIC:174675.178068, DBLP:conf/popl/LiangHJ95,
  DBLP:journals/pacmpl/DaraisLNH17, Sergey:2013:MAI:2491956.2491979}, by plugging
  different monads, the underlying semantics of the interpreter is changed.
  \todo{mention that direct-style also works?}
\item Env-store-value abstractions, which control whether the interpreter
  behaves as a concrete interpreter or an abstract interpreter.
  The environments, stores and values declared in the generic interpreter are
  abstract types; we also define several abstract methods that manipulate
  environments or stores.
  As shown by the AAM approach \cite{DBLP:journals/jfp/HornM12,
  DBLP:conf/icfp/HornM10}, if we widen the values to abstract domains (such as
  intervals) and refactor the store to a mapping from finite addresses to sets of
  values, we can derive a computable and sound abstract interpreter.
\item Binding-time abstractions, which control whether the interpreter produces
  values or residual programs. The binding-times differentiate whether the
  data is known statically or dynamically. Unsurprisingly, such binding-time
  information can be encoded into higher-kinded types \todo{cite FTPE, stage poly}.
\end{itemize}

\paragraph{Abstraction without Regret}

We show that these well-known abstractions can be combined smoothly, and more
importantly the price is almost free. Through staging, the major overhead of
abstractions -- from the monadic layers -- is completely eliminated in the
generated code. However, we do not lose any advantages of this high-level
programming style (e.g., the extensibility); the meaning of the program also
remains the same, particularly, the abstract semantics that does the analysis
when instantiated as an abstract interpreter.

@@

Besides the intellectual merit shown in the confluence diagram, this paper
contributes to a practical problem in static analysis: \textit{can we speed up a
static analyzer without any intrusive changes and soundness compromises?} Static
analysis is known as a trade-off game between \textit{precision} and
\textit{efficiency}, we argue that applying meta-programming (multi-stage
programming, in particular) with abstract interpreters is an effective approach
to improve the latter one while the former one is untouched.
In the literature, specializing analysis with respect to a program by partial
evaluation \cite{damian1999partial, amtoft1999partial,
Boucher:1996:ACN:647473.727587, ashley:practical} is not a new, despite the fact
that most of the previous work are ad-hoc on a particular analysis or requiring
significant changes to the analyzer. For example, abstract compilation
\cite{Boucher:1996:ACN:647473.727587} requires the whole analyzer to be
rewritten as closure generation from. In this paper, we advocate a systematic
approach to meta-programming that achieves the performance goal with less
intrusive changes; in fact, a case study (Section~\ref{cs_ac}) comparing with
abstract compilation shows, by utilizing type-based stage annotations, the
analyzer program does not need any changes.

@@

Our next contribution is to enable modular analysis based on a whole-program
analyzer by meta-programming. Modern software are shipped with large libraries
\cite{toman_et_al:LIPIcs:2017:7121}, and analyzing programs with large libraries
is usually expensive or unnecessarily repeated for whole-program analyzing
algorithms. For example, it has been shown that analyzing a simple ``Hello
World'' program in Java depends on additional 3,000 classes in the library
\cite{DBLP:conf/oopsla/KulkarniMZN16}. A natural idea is to analyze libraries
and generate summaries, later we can plugin the necessary information into the
summaries and finish the analysis; therefore such summaries should be modular
and reusable. Meta-programming gives us a chance to readily tweak a
whole-program analyzing algorithm into an analyzer that works modularly and
produces summaries. Still, the idea is to specialize the abstract interpreter
that was used for whole-program analysis, but on a subset of the program, for
examples, a set of generic functions for list manipulation. Such specialization
works on lambda expressions and generates code, which is partial analysis
results and can be completed by running with the necessary arguments (for
example, an environment that closes the lambda term) at next stages. Multiple
partial analysis results also can be reused later and composed together as they
are just normal programs at the next stage. Through staging, we mechanically
obtain a modular analysis, even though the original algorithm is formulated as a
whole-program analysis.

\paragraph{The paper is organized as follows:}

\begin{itemize}
  \item We progressively present a unifying construction that naturally extends
    the first Futamura projection to abstract interpreters. Starting from a generic
    interface of the interpreter for a small functional language
    (Section~\ref{generic_if}), we first show the instantiation of concrete
    interpretation (Section~\ref{unstaged_conc}). Then we lift it to the staged
    version by replacing the binding-time types (Section~\ref{stagedinterp}); and to
    the abstract interpreter by applying abstraction to the environment, store, and
    values (Section~\ref{unstaged_abs}). Finally, we show the combination of the two
    transformations, dubbed \textit{staged abstract interpreters}, can be easily
    derived (Section~\ref{sai}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

### 2019-03-15 paper/20_bg.tex
some revision

In this section, we first describe the abstract syntax of the language we will
analyze, then present the generic interface shared between the four different
interpreters. We choose to use Scala to demonstrate the idea, but the idea of
utilizing monads and multi-stages also can be easily constructed in other functional
languages, such as MetaOCaml \cite{DBLP:conf/gpce/CalcagnoTHL03,
  DBLP:conf/flops/Kiselyov14} or Template Haskell \cite{Sheard:2002:TMH:636517.636528}.
We first instantiate the interpreter to the concrete one; then we
proceed in two directions: by staging it we obtain a staged concrete interpreter
(Section~\ref{stagedinterp}); by abstracting it we obtain an unstaged abstract
interpreter (Section~\ref{unstaged_abs}).

>>>>

In this section, we first describe the abstract syntax of the language we will
analyze, then present the generic interface shared between the four different
interpreters. We choose to use Scala to demonstrate the idea, but the idea of
utilizing monads and multi-stages also can be easily constructed in other functional
languages, such as MetaOCaml \cite{DBLP:conf/gpce/CalcagnoTHL03,
  DBLP:conf/flops/Kiselyov14} or Template Haskell \cite{Sheard:2002:TMH:636517.636528} \todo{cite the recent one}.
We first instantiate the interpreter to the concrete one; then we
proceed in two directions: by staging it we obtain a staged concrete interpreter
(Section~\ref{stagedinterp}); by abstracting it we obtain an unstaged abstract
interpreter (Section~\ref{unstaged_abs}).

### 2019-03-15 paper/21_bg_interp.tex
some revision

In Scala, thanks for its nature that combines functional programming and
object-oriented programming, we may use traits and objects to implement monads.
For example, the following trait @Monad@ declares the @pure@ and @flatMap@
operation. The trait itself takes a higher-kinded type @F[_]: * -> *@
which takes a type and returns a type. The method @pure@ promotes values of type
@A@ to values of type @F[A]@. The monadic @bind@ operation is usually called
@flatMap@ in Scala, which takes a monad-encapsulated value of type @F[A]@, a function of
@A => F[B]@ and returns values of type @F[B]@.

@@

\paragraph{Stage Polymorphism}

>>>>

In Scala, thanks for its nature that combines functional programming and
object-oriented programming, we may use traits and objects to implement monads.
For example, the following trait @Monad@ declares the @pure@ and @flatMap@
operation. The trait itself takes a higher-kinded type @F[_]: * -> *@ as argument,
which takes a type and returns a type. The method @pure@ promotes values of type
@A@ to values of type @F[A]@ \footnote{We elect to use \texttt{pure} as the name, since \texttt{return} is a keyword in Scala and \texttt{unit} is a built-in function in LMS.}. The monadic @bind@ operation is usually called
@flatMap@ in Scala, which takes a monad-encapsulated value of type @F[A]@, a function of
@A => F[B]@ and returns values of type @F[B]@.

@@

\paragraph{Binding-time Abstraction}
%\paragraph{Stage Polymorphism}
% Binding-time abstraction via abstract types
